---
title: "a3_task3_thomas_lenihan"
author: "Thomas Lenihan"
date: "10/23/2021"
output: html_document
---
## Setup 

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) 
library(here)
library(janitor) 
library(ggplot2)
library(kableExtra)#attaching packages, setting up code chunk options

```

## Read in the data

```{r}
urchins <- read_csv(here("data", "urchins.csv")) %>% 
  clean_names() #put titles in lower snake case

```

## Data visualization

### Size Distribution Histogram
Comparing the distribution of sizes between purple and red urchins
```{r}
ggplot(data = urchins,
       aes(x = size)) +
  geom_histogram(binwidth = 1, color = "black", fill = "white") + 
  # create a histogram with binwidth 1, to show the pattern of size distribution
  facet_grid(cols = vars(common_name))+ # break up size distributions by species (common+_name)
  labs(title = "Urchin Size Distributions By Species",
       x = "Size (cm)",
       y = "frequency") #add axis titles and labels

```
*Figure 1. Distribution of Urchin Sizes by Species.* This figure shows the differences in size distribution across populations of Purple and Red Urchins.Sizes are the diameter of each urchin in centimeters, and are rounded to the nearest 0.5 cm. 

### Urchin Size Statistics Table

*Table 1. Descriptive Statistics of Urchin Sizes.* Sizes of Purple and Red Urchins were measured as the diameter of each individual in centimeters. mean, median, standard deviation, and sample size for each species are shown below. 
```{r}

urchin_size_stats <- urchins %>% #create a new data frame which will become the stats table
  group_by(common_name) %>% #organize by spp
  summarize(mean_mass_g = mean(size, na.rm = TRUE),
            median = median(size, na.rm = TRUE),
            sd = sd(size, na.rm = TRUE),
            sample_size = n()) #create summary statistics 

urchin_size_stats %>% 
  kable(col.names = c("Urchin Species", 
                      "Mean size (cm)", 
                      "Median size (cm)",
                      "Standard Deviation (cm)",
                      "Sample size (n)")) %>% 
  kable_styling(full_width = FALSE)
#create a table to show summary stats using kableExtra
```


```{r}

```

